---
title: "Montoya_Lab_Statistics"
author: "Anh H. Pham"
date: "2024-02-09"
output:
  html_document:
    df_print: paged
---


Clean up the environment and load necessary libraries.
```{r setup}
# Clean up environment
rm(list=ls())

# Libraries
require(readxl)
require(rstatix)
require(ggplot2)
require(dplyr)
```

Load the xlsx data
```{r}
df <- read_xlsx(file.choose()) %>% as.data.frame() # Navigate to data file
head(df, 10) # Print first 10 rows of table
```

Define the variable and group being tested
```{r}
group <- 'Cruise' # Change this
variable <- 'SST' # Change this
formula <- as.formula(paste(variable,'~',group)) # Create formula for stats tests
dec <- 2 # Rounding result to dec number of decimal points
```

Checking for statistical tests assumption If no assumption is violated, do parametric test (Student's T-test, ANOVA) If assumptions are violated, consider non-parametric test (Welch's T-Test, Welch's ANOVA, Kruskal-Wallis test)
```{r, warning=F}
# Shapiro's test for normal distribution
# p-values < 0.05: Data of group doesn't follow normal distribution
shapiro <- df %>% group_by(!!sym(group)) %>% shapiro_test(!!sym(variable)) 
# Use !!sym(variable) to use the variable dyanmically
print(shapiro)

# Histogram and density plot to visuallize data distribution among groups
ggplot(df, aes(x = .data[[variable]])) +
  geom_histogram(aes(y = ..density..), binwidth = 0.1, color = 'black') +
  geom_density(col = 'red') +
  facet_wrap(~ .data[[group]], scales = 'free_x')

# Levene's test for equal variance
# p-value < 0.05: Data among group doesn't have equal variance
levene <- levene_test(df,formula)
print(levene)

# Boxplot is good for checking variances
ggplot(df, aes(x = .data[[group]], y = .data[[variable]])) +
  geom_boxplot()

```

Create table of summary statistics
```{r}
summary_stats <- df %>% group_by(!!sym(group)) %>%
  summarise(
    'n' = n(),
    'Mean' = round(mean(!!sym(variable), na.rm = T),dec),
    'SD' = round(sd(!!sym(variable), na.rm = T),dec),
    'Mean ± SD' = paste(Mean, '±', SD),
    'Min' = round(min(!!sym(variable), na.rm = T), dec),
    'Max' = round(max(!!sym(variable), na.rm = T), dec),
    'Median' = round(median(!!sym(variable), na.rm = T),dec),
    '25th_quantile' = round(quantile(!!sym(variable), probs = 0.25, na.rm = T), dec),
    '75th_quantile' = round(quantile(!!sym(variable), probs = 0.75, na.rm = T), dec),
    'IQR' = round(IQR(!!sym(variable), na.rm = TRUE), dec)
  ) %>% as.data.frame() 
print(summary_stats)
```

ANOVA test to compare means of 3 or more groups Only use when no assumption was violated
If p-value < 0.05: There is a significant difference in the means of the variable among the groups being tested.
Tukey-HSD for pairwise comparison, only use when ANOVA p-value lower than 0.05
```{r, warning=F}
# ANOVA
anova <- anova_test(df,formula)
print(anova)

# Posthoc test
tukey <- tukey_hsd(df,formula)
print(tukey)

# Bar plot, note that this is base on the summary_stats table
ggplot(summary_stats,aes(x = .data[[group]], y = Mean, fill = .data[[group]]))+
  geom_bar(stat = 'identity')+
  geom_errorbar(aes(ymin=Mean-SD,ymax=Mean+SD),width=0.2)+
  theme_classic()
```

Kruskal-Wallis to compare distrution of 3 or more group Use when an assumption is violated.
If p-value < 0.05: There is a significant difference in the distribution of the variable among the groups being tested.
Dunn's test for pairwise comparison, Only use when KW p-value lower than 0.05
```{r}
# Kruskal-Wallis
kruskal <- kruskal_test(df,formula)
print(kruskal)

# Posthoc test
dunn <- dunn_test(df,formula)
print(dunn)

# Boxplot, note that this is based on the df table
ggplot(df,aes(x = .data[[group]], y = .data[[variable]], fill = .data[[group]]))+
  geom_boxplot()+
  theme_classic()
```

T-test example
Options:
* equal variance (Student's T-test) or not (Welch's T-Test)
* one-sided or two-sided
* paired or unpaired
```{r}
# Filter for only 2 group
ds <- df %>% filter(!!sym(group) %in% c('KN197','EN614')) # Change what's in c()
ttest <- t_test(ds, formula, var.equal=T, alternative="two.sided", paired=FALSE) # Change options
print(ttest)
```

Linear Regression
```{r}
x = 'MLD'
y = 'ChlMD'
formula = as.formula(paste(y,'~',x))
### Linear regression
model <- lm(formula,df)
summary(model)
# Extract and print the model equation
coefficients <- coef(model) %>% round(2)
cat("Model function: y =", coefficients[2], "*x + ", coefficients[1], "\n")
# Extract and print R-squared value
r_squared <- summary(model)$r.squared %>% round(2)
cat("R-squared value:", r_squared, "\n")

# Scatter plot
ggplot(df,aes(x=.data[[x]],y=.data[[y]]))+
  geom_point()+
  geom_smooth(method='lm',formula=y~x)
```
```